Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-11-14 papers/histogram/sad.tex

\begin{abstract}
  \jpsays{edit to include LJ clusters}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  the 31-atom Lennard-Jones cluster and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interest, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=1/3$ and $T=\infty$.}
  \label{fig:entropy-cartoon}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{gamma-n256}
  \caption{The update factor $\gamma_t$ versus iteration for $N=256$
    square-well atoms, for four different methods: WL, $1/t$-WL, SAMC,
    and SAD.}
    \label{fig:gamma-vs-t}
\end{figure}
Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for the related
methods SAD, WL, $1/t$-WL, and SAMC.  For SAMC, $\gamma_t$ remains
constant before dropping as $1/t$.  WL $\gamma_t$ remains at 1 for
many iterations, and then decreases very rapidly, with $1/t$-WL
behaving similarly but decreasing more aggressively before
transitioning to a more conservative $1/t$ behavior.  The update
factor for SAD fluctuates dynamically around a value less than 1 for
early MC moves, and then decreases as approximately $1/t$ while
continuing to fluctuate as new energies are found to be important.  At
intermediate times, the SAD $\gamma_t$ decreases as $1/t^2$ before
asymptoting to $N_S/t$, which is the same as $1/t$-WL.

@@

\subsection{Square-well fluid}
\davidsays{Edit this paragraph: introduce novelty of square well fluid
in particular, and importance.}
As our first test case, we consider the square-well fluid i.e. a
system of particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.  The
potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

@@

\subsubsection{50 square-well atoms}
\davidsays{check this periodic}
For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from $-248$ to $-120$.  The number of important energy states
for this system is therefore $N_S = 129$.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

@@

\subsubsection{256 square-well atoms}
\davidsays{check this periodic}
Next we will introduce a considerably larger simulation containing 256
atoms which has a maximum entropy about 1500 greater than its minimum.
This makes exploring the entire range of energies extremely expensive,
and strongly favors the methods that restrict the energy (or
temperature) range of interest.  For this simulation, we chose a much
higher minimum reduced temperature of $1.0$, which corresponds to an
interesting energy range from $-915$ to $-509$.  The number of
important energy states for this system is then $N_S = 407$.  The
minimum entropy over this energy range is just 395 less than the
maximum, corresponding to a ratio of only $10^{118}$ between the
highest and lowest density of states.

@@

In order to explore both the bin-size behavior and low-temperature
effectiveness of our algorithm, we chose to study the 31-atom cluster
(LJ31).
Clusters of Lennard-Jones atoms have been frequently used for testing
Monte Carlo algorithms~\davidsays{add a few citations}.  The LJ31
cluster is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\davidsays{add citations}.
In particular, the LJ31 cluster features a solid-solid phase
transition at a temperature of around $T=0.03\epsilon/k_B$ that is
challenging to converge~\davidsays{cite}.  Poulain \emph{et al.}
tested a variety of variants of WL on LJ31 and found
\davidsays{summarize}~\cite{poulain2006performances}.

@@

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster as
a function of temperature.  There are two peaks, a large peak at 0.2?
corresponding to melting temperature, and a small peak around 0.03 for
the solid-solid transition from a mackay to antimackay
wahatever~\davidsays{cite}.  We keep our atoms in a spherical box with
radius 2.5$\sigma$, as others do~\davidsays{cite}.  We find that using
a larger box (with radius $5\sigma$) has a large effect on the melting
peak, and a smaller but still significant effect on the solid-solid
transition.  Since our focus is on the convergence rate, we focused on
the smaller box size.  We further restricted our simulations to
negative energies, as the microcanonical energy at infinite
temperature is rather high for such a small box.

@@

The heat capacity changes significantly even at these low temperatures
when the size of the cavity is increased.  We use a 2.5 cavity because
``everyone else'' does~\cite{poulain2006performances}.  For purposes
of testing the convergence of the algorithm, we don't need to solve
for the true heat capacity.  Furthermore, the heat capacity at fixed
volume is not the most meaningful quantity to predict for a cluster.

>>>>

\begin{abstract}
  \jpsays{edit to include LJ clusters when paper finished!}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  the 31-atom Lennard-Jones cluster and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interest, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{N50-lndos-comparison}
  \caption{The entropy of a square-well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=1/3$ and $T=\infty$.}
  \label{fig:entropy-cartoon}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{gamma-n256}
  \caption{The update factor $\gamma_t$ versus iteration for the
    square-well fluid with 256 atoms, for four different methods: WL, $1/t$-WL, SAMC,
    and SAD.}
    \label{fig:gamma-vs-t}
\end{figure}
Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for the related
methods SAD, WL, $1/t$-WL, and SAMC.  For SAMC, $\gamma_t$ remains
constant before dropping as $1/t$.  WL $\gamma_t$ remains at 1 for
many iterations, and then decreases very rapidly, with $1/t$-WL
behaving similarly but decreasing more aggressively before
transitioning to a more conservative $1/t$ behavior.  The update
factor for SAD fluctuates dynamically around a value less than 1 for
early MC moves, and then decreases as approximately $1/t$ while
continuing to fluctuate as new energies are found to be important.  At
intermediate times, the SAD $\gamma_t$ decreases as $1/t^2$ before
asymptoting to $N_S/t$, which is the same as $1/t$-WL.

@@

\subsection{Square-well fluid}
As our first test case, we consider the square-well fluid i.e. a
system of particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.
\jpcom{In addition to providing an ideal test-bed for new algorithm development,
to the best of the knowledge of the authors, there are \emph{not} any direct convergence comparison tests in the literature for the variety of flat histogram methods presented here applied to the square-well fluid.}
The
potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

@@

\subsubsection{50 square-well atoms}
For this 50 atom simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from $-248$ to $-120$.  The number of important energy states
for this system is therefore $N_S = 129$.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

@@

\subsubsection{256 square-well atoms}
Next we introduce a considerably larger simulation containing 256
atoms which has a maximum entropy about 1500 greater than its minimum.
This makes exploring the entire range of energies extremely expensive,
and strongly favors the methods that restrict the energy (or
temperature) range of interest.  For this simulation, we chose a much
higher minimum reduced temperature of $1.0$, which corresponds to an
interesting energy range from $-915$ to $-509$.  The number of
important energy states for this system is then $N_S = 407$.  The
minimum entropy over this energy range is just 395 less than the
maximum, corresponding to a ratio of only $10^{118}$ between the
highest and lowest density of states.

@@

In order to explore both the bin-size behavior and low-temperature
effectiveness of our algorithm, we chose to study the 31-atom cluster
(LJ31) because it is the smallest Lennard-Jones cluster that exhibits broken ergoidcity
and a solid-solid phase transition at a low temperature~\cite{doye1998thermodynamics, martiniani2014superposition}.
Clusters of Lennard-Jones atoms have been frequently used for testing
Monte Carlo algorithms~\cite{wales1997global, neirotti2000phase, frantsuzov2005size, mandelshtam2006multiple}.  The LJ31
cluster is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\cite{calvo2000entropic, calvo2000phase, poulain2006performances, martiniani2014superposition}.
In particular, the LJ31 cluster features a solid-solid phase
transition at a temperature of around $T \approx 0.027\epsilon/k_B$ that is
challenging to converge~\cite{mandelshtam2006multiple}.  Poulain \emph{et al.}
tested a number of variants of WL on LJ31 and found
\davidsays{summarize}~\cite{poulain2006performances}.

@@

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster as a function
of temperature.  There are two peaks, a large peak around $T \approx
0.3\epsilon/k_B$? corresponding to melting temperature, and a small peak around
$T \approx 0.027\epsilon/k_B$ for the solid-solid transition from a
Mackay-to-anti-Mackay (M $\rightarrow$ aM) transition~\cite{doye1997thermally,
hendy2001molecular, mandelshtam2006multiple}.  We keep our atoms in a spherical
box with radius 2.5$\sigma$, as is common in the
literature~\cite{poulain2006performances, mandelshtam2006multiple}.  We find
that using a larger box (with radius $5\sigma$) has a large effect on the
melting peak, and a smaller but still significant effect on the solid-solid
transition.  Since our focus is on the convergence rate, we focused on the
smaller box size.  We further restricted our simulations to negative energies,
as the microcanonical energy at infinite temperature is rather high for such a
small box.

@@

\jpcom{this is a basically a repeat of above...

The heat capacity changes significantly even at these low temperatures
when the size of the cavity is increased.  We use a 2.5 cavity because
``everyone else'' does~\cite{poulain2006performances}.  For purposes
of testing the convergence of the algorithm, we don't need to solve
for the true heat capacity.  Furthermore, the heat capacity at fixed
volume is not the most meaningful quantity to predict for a cluster.
}
